# Reproducible Research: Peer Assessment 1
Title:  PA1_template.Rmd                     ##  
Author: Ambika J                             ##  
Date:   Jul-2014                             ##  
Notes:  Reproducible Research Assignmnt 1    ##  

## Loading and preprocessing the data
```{r filedownload,echo=FALSE,results=FALSE}
setwd("/Users/ambika/gitDir/RepData_PeerAssessment1")
if(!file.exists("activity.csv")) unzip("activity.zip")
```

```{r loadData}
activity <- read.csv("activity.csv", colClasses = c("integer","character","integer"), comment="",stringsAsFactors = FALSE, nrows = 17568)
```

## What is mean total number of steps taken per day?
```{r stepsMean}
library(reshape2)

newactivity <- activity[complete.cases(activity$steps),]
dt <- factor(as.Date(newactivity$date))
totalsteps <- melt(tapply(newactivity$steps, dt, sum))
dimnames(totalsteps)[[2]] <- c("date","total")
mn1 <- mean(totalsteps$total)
md1 <- median(totalsteps$total)
```

Total number of steps taken each day
```{r plotHist}
hist(totalsteps$total,breaks=50,col="lightblue",xlab='Total Steps', main = "Total number of steps taken each day")
```

Mean is `r sprintf("%.0f",mn1) `. Median is `r sprintf("%.0f",md1)`.

## What is the average daily activity pattern?
```{r dailyPattern}
library(reshape2)

sumsteps <- melt(tapply(newactivity$steps,  newactivity$interval, sum))
countdates <- melt(tapply(newactivity$date, newactivity$interval, length))
merged <- merge(sumsteps,countdates, by.x = "Var1", by.y = "Var1")
averaged <- cbind(merged,merged$value.x/merged$value.y)
dimnames(averaged)[[2]] <- c("interval","totalsteps","countdates","average")

max5min <- averaged$interval[which.max(averaged$average)]
```

Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r plotL}
plot(averaged$average ~ averaged$interval, type = "l", xlab = "5-minute interval", ylab = "average steps",col="blue")
```

`r max5min` is the 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.

## Imputing missing values
```{r missingValues}
sum(is.na(activity$steps))
```
`r sum(is.na(activity$steps))` is the total number of missing values [steps] in the dataset.

A new data set with the new strategy to fill in the missing values. Here, the NA values in step is filled with the median value for each interval.
```{r fillStrategy}
medsteps <- melt(tapply(newactivity$steps,  newactivity$interval, median))
filledData <- merge(activity,medsteps, by.x = "interval", by.y = "Var1",all = TRUE)
filledData[["steps"]][is.na(filledData[["steps"]])] <- filledData[["value"]][is.na(filledData[["steps"]])]
reorderFilledData <- filledData[order(filledData$date,filledData$interval),c(2,3,1)]
#reorderFilledData[is.na(activity[["steps"]]),] 
#tapply(reorderFilledData[["steps"]][is.na(activity[["steps"]])], reorderFilledData$date, sum)

totalfilledsteps <- melt(tapply(reorderFilledData$steps, factor(as.Date(reorderFilledData$date)), sum))
dimnames(totalfilledsteps)[[2]] <- c("date","total")
mn2 <- mean(totalfilledsteps$total)
md2 <- median(totalfilledsteps$total)
```

Total number of steps taken each day (after filling missing values)
```{r plotFillHist}
hist(totalfilledsteps$total,breaks=50,col="lightblue",xlab='Total Steps', main = "Total number of steps taken each day (after filling missing values)")
```

After filling the missing values, mean is `r sprintf("%.0f",mn2) `. Median is `r sprintf("%.0f",md2)`.

Do these values differ from the estimates from the first part of the assignment? 
* YES, they do.  

What is the impact of imputing missing data on the estimates of the total daily number of steps?  
Following are the observations:  
* As median is considered to fill in the values, the mean and median has decreased by `r ((mean(totalsteps$total) - mean(totalfilledsteps$total)) / mean(totalsteps$total)) * 100`% and `r ((median(totalsteps$total) - median(totalfilledsteps$total)) / median(totalsteps$total)) * 100`% respectively.  
* Before and After:  
Before imputing missing data mean and median is `r sprintf("%.0f",mn1) ` and `r sprintf("%.0f",md1)` respectively.  
After imputting missing  data mean and median is `r sprintf("%.0f",mn2) ` and `r sprintf("%.0f",md2)` respectively.  
* As per histogram comparison, there are about 8 days increase in the range of 1000 to 2000 total steps; which is not seen in the first part of the assignment.  
* More specific details at date level [against total steps for each date]:  
```{r snippet,echo=FALSE}
tapply(reorderFilledData[["steps"]][is.na(activity[["steps"]])], reorderFilledData$date[is.na(activity[["steps"]])], sum)
```
 
## Are there differences in activity patterns between weekdays and weekends?
```{r weekday}
reorderFilledData$date <- as.Date(reorderFilledData$date)
weekend <- weekdays(as.Date(reorderFilledData$date),TRUE) %in% c("Sat", "Sun")
reorderFilledData$day <- factor(ifelse(weekend, "weekend", "weekday"))
sumfillsteps <- melt(tapply(reorderFilledData$steps,  reorderFilledData$interval, sum))
countdays <- melt(tapply(reorderFilledData$day, reorderFilledData$interval, length))
mergedfill <- merge(sumfillsteps,countdays, by.x = "Var1", by.y = "Var1")
averagedfill <- cbind(mergedfill,mergedfill$value.x/mergedfill$value.y)
dimnames(averagedfill)[[2]] <- c("interval","totalsteps","countdays","average")
```
```{r weeksplot}
library(lattice)
xyplot(average ~ interval | countdays, data = averagedfill, layout = c(1, 2), type="l")
```
```{r cleanup, echo=FALSE}
rm(list=ls())
```